---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}

#Load data that is saved in the Working Directory

activity<-read.csv("activity.csv",stringsAsFactors=FALSE)

#Transforming date from character and integer variables to numeric for ease of calculations
activity$date<-as.Date(activity$date)
activity$steps<-as.numeric(activity$steps)
activity$interval<-as.numeric(activity$interval)


```

## What is mean total number of steps taken per day?

```{r fig.height=5}

#Calculating total number of steps per day

# Removing the Missing values while calculating total, as per instructions missings to be ignored

library(dplyr)

StepsPerDay<-summarise(group_by(activity,date),
                       totalStepsPerDay=sum(steps,na.rm=TRUE))

# Histogram of the total number of steps taken each day

hist(StepsPerDay$totalStepsPerDay,col="Red",main='Total Steps Per Day',
     xlab='Steps Per Day',ylab='Frequency' )

#Mean and MEdian of Total Steps per Day,  Removing the Missing values while calculating mean/median
meanStepsPerDay<- mean(StepsPerDay$totalStepsPerDay,na.rm=TRUE)
medianStepsPerDay<- median(StepsPerDay$totalStepsPerDay,na.rm=TRUE)

meanStepsPerDay
medianStepsPerDay
```

## What is the average daily activity pattern?

```{r}
#No. of steps per 5-minute interval across all days
meanStepsPerInterval<-summarise(group_by(activity,interval),
                            meanStepsPerInt=mean(steps,na.rm=TRUE))

# Plotting the Average steps per interval
with(meanStepsPerInterval,plot(interval,
                               meanStepsPerInt,
                               type='l',
                               ylab='Average steps per Interval',
                               xlab='Intervals'))

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

subset(meanStepsPerInterval$interval,meanStepsPerInterval$meanStepsPerInt==max(meanStepsPerInterval$meanStepsPerInt))

#08:35 is the max interval
```

## Imputing missing values
```{r}
#NAs in the dataset
sum(is.na(activity))

#Merging the average steps per interval from a previous step
activity2<-merge(x = activity,    y = meanStepsPerInterval,    by = "interval", all.x=TRUE)
```
###Missing Values are imputed using the Mean valus of the steps computed for each interval across all days
```{r}
#Imputing the missing values in steps with the mean steaps per interval 
activity2$steps_meanimputed<-ifelse(is.na(activity2$steps),activity2$meanStepsPerInt,activity2$steps)

#Create new dataset replacing the old steps column with mean-imputed steps column
activity2<-activity2[,c(5,3,1)]
names(activity2)<-c("steps","date","interval")

#Histogram after the mean imputation

StepsPerDay2<-summarise(group_by(activity2,date),
                       totalStepsPerDay2=sum(steps,na.rm=TRUE))
head(StepsPerDay2)
summary(StepsPerDay2)

hist(StepsPerDay2$totalStepsPerDay2,col="Green",main='Total Steps Per Day',
     xlab='Steps Per Day',ylab='Frequency' )

#Mean and MEdian of Total Steps per Day
meanStepsPerDay2<- mean(StepsPerDay2$totalStepsPerDay2,na.rm=TRUE)
medianStepsPerDay2<- median(StepsPerDay2$totalStepsPerDay2,na.rm=TRUE)

meanStepsPerDay2
medianStepsPerDay2


```
###The Mean and Median changes slightly after the imputation, however the changes are small as imputation is done using Mean

## Are there differences in activity patterns between weekdays and weekends?
```{r}

library(ggplot2)
#Weekdays/Weekend analyses

activity2$wkday<-weekdays(activity2$date)

#weekday or weekday
activity2$daytype<-as.factor(ifelse(activity2$wkday %in% c("Saturday","Sunday"),"weekend","weekday"))

AvgStepsPerInterval<-summarise(group_by(activity2,daytype,interval),
                        AvgStepsPerInt=mean(steps,na.rm=TRUE))

f1<-ggplot(data=AvgStepsPerInterval,aes(x=interval,y=AvgStepsPerInt)) + geom_line()
f2<-f1+facet_grid(daytype~.)
f3<-f2+labs(x="Interval",y="Average Steps by Interval",title="Comparing Average Steps per Interval by Weeday/Weekend")

f3

```
###Higher averages in the Weekends are more generally but mornings are higher for weekdays